Particle Smoothing Algorithms for Variable Rate Models

### 2012-04-25 full_finaldraft_VRsmoothing_withSJGcomments.tex
Done Simon's suggestions.

The objective for inference will be to estimate the changepoint sequence. This will be denoted as $\theta = \{\tau_{0:K}, u_{0:K}\}$. At a particular time $t_n$, the sequence will be divided into past $\theta_n = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$. It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$.

@@

\begin{IEEEeqnarray}{rCl}
p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
\end{IEEEeqnarray}

The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}. The number of changepoints, $K_n$, is not fixed. If $u_k$ varies over $\mathcal{U}$, and $\tau_k$ varies over the real line $\mathbb{R}$, then the space spanned by $\theta_n$ is the union,
%
\begin{IEEEeqnarray}{rCl}
 \Theta & = & \bigcup_k \Theta_k \label{eq:theta_space} \\
 \Theta_k & = & \{k\} \times \mathbb{R}^k \times \mathcal{U}^k     .
\end{IEEEeqnarray}

@@

We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM,
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n})x_n + v_n       ,
\end{IEEEeqnarray}

where $w_n$ and $v_n$ are independent Gaussian variables (although the independence condition may be relaxed \cite{Grewal2002}),
%
\begin{IEEEeqnarray}{rCl}
 w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n})) \\
 v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}))       .
\end{IEEEeqnarray}

@@

Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems. For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated in an optimal fashion.

@@

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$, which is defined on the same space as the prior density (\ref{eq:theta_space}). This distribution may be expanded using Bayes' rule,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}|y_{1:n}) \propto p(y_n|\theta_{n}, y_{1:n-1}) } \nonumber \\
 \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}     .
\end{IEEEeqnarray}

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution. In this case, each particle will be a set of changepoint times and parameters.
%
\begin{equation}
 \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n}) \label{eq:vrpf}
\end{equation}

where $\delta_x(X)$ is a unit probability mass at $X=x$.

The particle filter works recursively. At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights, $\{v_{n-1}^{(j)}\}$ (where $\sum_j v_{n-1}^{(j)} = 1$),
%
\begin{equation}
 q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})     .
\end{equation}

The choice of weights determines the type of resampling used. The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}|y_{1:n-1})$. This, however, leads to degeneracy of the particle weights over time. Conventional resampling is achieved by using $v_{n-1}^{(j)} = w_{n-1}^{(j)}$. Any other choice results in an auxiliary particle filter \cite{Pitt1999}. For further discussion of resampling, see \cite{Cappe2007,Doucet2009,Douc2005}.

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}^{(i)}$ to create a proposal for $\theta_n^{(i)}$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}^{(i)}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) p(\theta_{n-1}^{(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

It only remains to consider the likelihood term required for evaluation of the importance weights, $p(y_n|\theta_n, y_{1:n-1})$. The form of this term depends on the particular model under consideration. In the following sections, the likelihood expressions for the conditionally linear-Gaussian and deterministic cases are considered.

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

with means and variances given by the following standard recursions (dependence on $\theta_{n}$ suppressed for clarity),
%
\begin{IEEEeqnarray}{rCl}
 m_n^- & = & A_n m_{n-1} \label{eq:kf_predict_start} \\
 P_n^- & = & A_n P_{n-1} A_n^T + Q_n \\
 \mu_n & = & C_n m_n^- \\
 S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict_stop} \\
 K_n   & = & P_n^- C_n^T S_n^{-1} \label{eq:kf_update_start}\\
 m_n   & = & m_n^- + K_n (y_n - \mu_n) \\
 P_n   & = & P_n^- - K_n S_n K_n^T \label{eq:kf_update_stop}    .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

When the conditionally deterministic models of section~\ref{sec:cd_models} are used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}). Thus, the required likelihood term is simply given by,
%
\begin{IEEEeqnarray}{rCl}
 p(y_n|\theta_{n}, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the target distribution may be factorised as follows at time $t_n$,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}|\theta_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

This suggests a sequential strategy for particle smoothing. Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}$ by simply discarding the changepoints before $t_n$. New values of $\theta_{n}$ are then sampled from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated. By iterating backwards over $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.

It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$. The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e. $p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$. However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\theta_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}, \theta_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})     .
\end{IEEEeqnarray}

The particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$, yielding,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n}) \label{eq:rb-vrps_back_cond}     ,
\end{IEEEeqnarray}

where the backwards conditional weights are given by,
%
\begin{IEEEeqnarray}{rCl}
 \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}     .
\end{IEEEeqnarray}

>>>>

The objective for inference will be to estimate the changepoint sequence. This will be denoted as $\theta = \{\tau_{0:K}, u_{0:K}\}$. At a particular time $t_n$, the sequence will be divided into past $\theta_n^- = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$. It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$.

@@

\begin{IEEEeqnarray}{rCl}
p(\theta_n^-) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
\end{IEEEeqnarray}

The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}. The number of changepoints, $K_n$, is not fixed. If $u_k$ varies over $\mathcal{U}$, and $\tau_k$ varies over the real line $\mathbb{R}$, then the space spanned by $\theta_n^-$ is the union,
%
\begin{IEEEeqnarray}{rCl}
 \Theta & = & \bigcup_k \Theta_k \label{eq:theta_space} \\
 \Theta_k & = & \{k\} \times \mathbb{R}^k \times \mathcal{U}^k     .
\end{IEEEeqnarray}

@@

We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM,
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n}^-)x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n}^-)x_n + v_n       ,
\end{IEEEeqnarray}

where $w_n$ and $v_n$ are independent Gaussian variables (although the independence condition may be relaxed \cite{Grewal2002}),
%
\begin{IEEEeqnarray}{rCl}
 w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n}^-)) \\
 v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}^-))       .
\end{IEEEeqnarray}

@@

Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems. For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n^-)$, is inflated in an optimal fashion.

@@

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}^-| y_{1:n})$, at each time $t_n$, which is defined on the same space as the prior density (\ref{eq:theta_space}). This distribution may be expanded using Bayes' rule,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}^-|y_{1:n}) \propto p(y_n|\theta_{n}^-, y_{1:n-1}) } \nonumber \\
 \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}^-) p(\theta_{n-1}^-|y_{1:n-1}) \label{eq:vrpf_target}     .
\end{IEEEeqnarray}

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution using importance sampling (IS). In this case, each particle will be a set of changepoint times and parameters.
%
\begin{equation}
 \hat{p}(\theta_{n}^-|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-) \label{eq:vrpf}
\end{equation}

where $\delta_x(X)$ is a unit probability mass at $X=x$. The particle filter works recursively. At the $n$th step, a particle, $\theta_{n-1}^{-(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights, $\{v_{n-1}^{(j)}\}$ (where $\sum_j v_{n-1}^{(j)} = 1$),
%
\begin{equation}
 q(\theta_{n-1}^-) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)     .
\end{equation}

The choice of weights determines the type of resampling used. The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}^-|y_{1:n-1})$. This, however, leads to degeneracy of the particle weights over time. Conventional resampling is achieved by using $v_{n-1}^{(j)} = w_{n-1}^{(j)}$. Any other choice results in an auxiliary particle filter \cite{Pitt1999}. For further discussion of resampling, see \cite{Cappe2007,Doucet2009,Douc2005}.

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_n)$, and concatenated with $\theta_{n-1}^{-(i)}$ to create a proposal for $\theta_n^-{(i)}$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{-(i)}|y_{1:n}) }{ q(\theta_{n}^{-(i)}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) p(\theta_{n-1}^{-(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{-(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

It only remains to consider the likelihood term required for evaluation of the importance weights, $p(y_n|\theta_n^-, y_{1:n-1})$. The form of this term depends on the particular model under consideration. In the following sections, the likelihood expressions for the conditionally linear-Gaussian and deterministic cases are considered.

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n^-, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}^-, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

with means and variances given by the following standard recursions (dependence on $\theta_{n}^-$ suppressed for clarity),
%
\begin{IEEEeqnarray}{rCl}
 m_n^- & = & A_n m_{n-1} \label{eq:kf_predict_start} \\
 P_n^- & = & A_n P_{n-1} A_n^T + Q_n \\
 \mu_n & = & C_n m_n^- \\
 S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict_stop} \\
 K_n   & = & P_n^- C_n^T S_n^{-1} \label{eq:kf_update_start}\\
 m_n   & = & m_n^- + K_n (y_n - \mu_n) \\
 P_n   & = & P_n^- - K_n S_n K_n^T \label{eq:kf_update_stop}    .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{-(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{-(i)})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

When the conditionally deterministic models of section~\ref{sec:cd_models} are used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n^-$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}). Thus, the required likelihood term is simply given by,
%
\begin{IEEEeqnarray}{rCl}
 p(y_n|\theta_{n}^-, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the target distribution may be factorised as follows at time $t_n$,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

This suggests a sequential strategy for particle smoothing. Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}^-$ by simply discarding the changepoints before $t_n$. New values of $\theta_{n}^-$ are then sampled from the conditional distribution, $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated. By iterating backwards over $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.

It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})$. The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e. $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}^-|\theta_{n}^{+}, y_{1:n}))$. However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}^-|\theta_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}^-, \theta_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}^-, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}^-, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\theta_{n}^+|\theta_{n}^-) p(\theta_{n}^-|y_{1:n})     .
\end{IEEEeqnarray}

The particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}^-|y_{1:n})$, yielding,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}^-|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{-(i)}}(\theta_{n}^-) \label{eq:rb-vrps_back_cond}     ,
\end{IEEEeqnarray}

where the backwards conditional weights are given by,
%
\begin{IEEEeqnarray}{rCl}
 \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
             &         & \times p(x_n|\theta_{n}^{-(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{-(i)}) \label{eq:rb-vrps_back_cond_weight}     .
\end{IEEEeqnarray}
